---
title: "MiceExtVal example of use"
author: 
 - name: "Asier Ballesteros Domínguez"
   orcid: 0000-0002-2195-4669
output: html_document
toc: true
bibliography: references.bib
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{gtsummary}
  %\VignetteDepends{kableExtra}
---

# Required libraries

The following code snippet lists all the libraries needed to run this report.

```{r}
#| message: false
#| warning: false
#| code-fold: false
library(MiceExtVal)
library(dplyr)
library(tidyr)
library(mice)
library(kableExtra)
library(gtsummary)
library(here)
library(tidyr)
library(tibble)
library(survival)
```

As an example of the package `MiceExtVal` use we will externally validate the Framingham model, see @framingham. We will use a dataset obtained from [kaggle](https://www.kaggle.com). The example of use will be structured as follows.

1. Read and explore the dataset
2. Impute the missing data
3. Define the Framingham model
4. Externally validate the model

# Read and Explore the Dataset

To recreate an external validation of the Framingham model we need a dataset with the model variables. In the package we have included the Framingham dataset provided in [kaggle](https://www.kaggle.com) by [Shrey Jain](https://www.kaggle.com/datasets/shreyjain601/framingham-heart-study). The dataset contains the variables described in next table. We can observe that there exists some missing values in some variables. 

```{r, warning=FALSE, echo=FALSE}
#| warning: false
#| label: tbl-summary-data
#| tbl-cap: Summary table of the Framingham dataset
#| classes: plain .table-sm
#| code-fold: true

MiceExtVal::framingham |>
  dplyr::select(-randid, -period) |>
  gtsummary::tbl_summary(
    data = _,
    by = anychd,
    statistic = list(
      gtsummary::all_continuous() ~ "{mean} ({sd})",
      gtsummary::all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    digits = gtsummary::all_continuous() ~ 2,
    missing_text = "(Missing)"
  )
```

In the following table there are represented the percentage o missings in the variables that have at least one row with a missing value. 

```{r, echo=FALSE}
#| echo: false
#| tbl-cap: Percentage of missings
#| label: tbl-perc-miss
#| classes: plain .table-sm
MiceExtVal::framingham |>
  dplyr::summarise(
    dplyr::across(dplyr::everything(), ~ mean(is.na(.)) * 100)
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::everything(),
    names_to = "variable",
    values_to = "pct_missing"
  ) |>
  dplyr::filter(pct_missing > 0) |>
  dplyr::arrange(dplyr::desc(pct_missing)) |>
  dplyr::rename(Variable = variable, `Percentage of Missings` = pct_missing) |>
  kableExtra::kbl(digits = 2, align = c("l", "c"))
```

# Impute Missing Data

To ammend these missing values we will impute their expected values using the mice package, see @mice-package. We will execute the default imputation as we are no interested in be so precise, we just want to reproduce an use case where we need to impute some missing data and from this data externally validate one or more models.

```{r}
#| eval: false
pred_matrix <- mice::make.predictorMatrix(MiceExtVal::framingham)
pred_matrix[, "randid"] <- 0
pred_matrix[, "period"] <- 0

mice_data <- mice::mice(MiceExtVal::framingham)
fram_long <- mice::complete(mice_data, "long")
```

```{r}
#| include: false
pred_matrix <- mice::make.predictorMatrix(MiceExtVal::framingham)
pred_matrix[, "randid"] <- 0
pred_matrix[, "period"] <- 0

mice_data <- mice::mice(MiceExtVal::framingham, predictorMatrix = pred_matrix)
fram_long <- mice::complete(mice_data, "long") |>
  dplyr::filter(period == 1) |>
  dplyr::rename(id = randid)
```

# The Framingham Model

With the data ready to be used we need to take some time to understand the model that we want to validate, the Framingham model. This model is a Cox model that predicts the risk to suffer a coronary heart disease in patients that are $30-79$ years old. The Framingham model coefficients are represented in the following table.

```{r}
#| echo: false
#| tbl-cap: Framingham model coefficients
#| tbl.alt: >
#|  Framingham model coefficients
#| label: tbl-framingham-coefficients
#| classes: plain .table-sm
options(knitr.kable.NA = "-")

tibble::tribble(
  ~Variable, ~Men, ~Women,
  "Age, y", 0.04826, 0.33766,
  "Age squared, y", NA, -0.00268,
  "Diabetes", 0.42839, 0.59626,
  "Smoker", 0.52336, 0.29246,
  "<160", -0.65945, -0.26138,
  "160-199", NA, NA,
  "200-239", 0.17692, 0.20771,
  "240-279", 0.50539, 0.24385,
  ">280", 0.65713, 0.53513,
  "<35", 0.49744, 0.84312,
  "35-44", 0.24310, 0.37796,
  "45-49", NA, 0.19785,
  "50-59", -0.05107, NA,
  ">60", -0.48660, -0.42951,
  "Optimal", -0.00226, -0.53363,
  "Normal", NA, NA,
  "High Normal", 0.28320, -0.06773,
  "Stage I hypertension", 0.52168, 0.26288,
  "Stage II-IV hypertension", 0.61859, 0.46573,
  "Baseline survival function at 10 years, S(t)", 0.90015, 0.96246
) |>
  kableExtra::kbl(digits = 4, align = c("l", "c", "c")) |>
  kableExtra::pack_rows("TC, mg/dL", 5, 9, bold = FALSE) |>
  kableExtra::pack_rows("HDL-C, mg/dL", 10, 14, bold = FALSE) |>
  kableExtra::pack_rows("Blood Pressure", 15, 19, bold = FALSE)
```

## Requirements to Define the Model

The Framingham model had defined the variables that are used to calculate the model predictions. In the dataset we have the data needed to meet this requirements. Before defining the model we need to generate the variables that are not present in the dataset. In the previous table we can observe that some continous variables are converted into categorical. More precisely, the total cholesterol, HDL cholesterol and blood pressure are categorized.

### Total Cholesterol

The total cholesterol variable is categorized into five categories, less than 160 mg/dL, between 160 and 199 mg/dL, between 200 and 239 mg/dL, between 240 and 279 mg/dl, and more than 280 mg/dL. The following code snippet generates the variable in the dataset.

| Variable             |   Men   |  Women  |
| :------------------- | :-----: | :-----: |
| **TC, mg/dL**        |         |         |
| <160                 | -0.6594 | -0.2614 |
| 160-199              |    -    |    -    |
| 200-239              | 0.1769  | 0.2077  |
| 240-279              | 0.5054  | 0.2439  |
| >280                 | 0.6571  | 0.5351  |

```{r}
fram_long <- fram_long |>
  dplyr::mutate(
    totchol_less_160 = as.numeric(totchol < 160),
    totchol_160_199 = as.numeric(totchol >= 160 & totchol < 200),
    totchol_200_239 = as.numeric(totchol >= 200 & totchol < 240),
    totchol_240_279 = as.numeric(totchol >= 240 & totchol < 280),
    totchol_greater_280 = as.numeric(totchol >= 280),
  )
```

### HDL cholesterol

The Framingham model characterizes the HDL cholesterol into five groups, less than 35 mg/dL, between 35 and 44 mg/dL, between 45 and 49 mg/dL, between 50 and 59 mg/dL, and more than 60 mg/dL. As before the following code snippet generates the variable in the dataset.

| Variable             |   Men    |  Women   |
| :------------------- | :------: | :------: |
| **HDL, mg/dL**       |          |          |
| <35                  | 0.49744  | 0.84312  |
| 35-44                | 0.24310  | 0.37796  |
| 45-49                |    -     | 0.19785  |
| 50-59                | -0.05107 |    -     |
| >60                  | -0.48660 | -0.42951 |
: {.plain .table-sm}

```{r}
fram_long <- fram_long |>
  dplyr::mutate(
    hdl_less_35 = as.numeric(hdlc < 35),
    hdl_35_44 = as.numeric(hdlc >= 35 & hdlc < 44),
    hdl_45_49 = as.numeric(hdlc >= 45 & hdlc < 49),
    hdl_50_59 = as.numeric(hdlc >= 50 & hdlc < 59),
    hdl_greater_60 = as.numeric(hdlc >= 60),
  )
```

### Blood pressure

In the model it is defined a variable that aggregates the systolyc blood pressure and diastolic blood pressure into five groups of hypertension. As defined by @framingham, we categorize each of the individual blood pressure variables into five groups, optimal, normal, high-normal, hypertension I, and hypertension II-IV. If a patient have different blood pressure categories we use the highest one as a representation of they state. The following code adds the variable to the dataset, we first generate the individual categorization variables and generate the `pressure` variable as the aggregation of both.

| Variable                 | Men      | Women    |
| :----------------------- | :------- | :------- |
| **Blood Pressure**       |          |          |
| Optimal                  | -0.00226 | -0.53363 |
| Normal                   | -        | -        |
| High Normal              | 0.28320  | -0.06773 |
| Stage I hypertension     | 0.52168  | 0.26288  |
| Stage II-IV hypertension | 0.61859  | 0.46573  |
: {.plain .table-sm}

We first generate a variable called `pressure` that categorize both pressures into one variable. 

```{r}
pressure_labels <- c(
  "optimal", "normal", "high-normal", "hypertension I", "hypertension II-IV"
)
fram_long <- fram_long |>
  dplyr::mutate(
    sys_lvl = cut(
      sysbp,
      breaks = c(-Inf, 120, 130, 140, 160, Inf),
      labels = pressure_labels
    ),
    dys_lvl = cut(
      diabp,
      breaks = c(-Inf, 80, 85, 90, 100, Inf),
      labels = pressure_labels
    ),
    pressure = purrr::map2_int(
      as.numeric(sys_lvl), as.numeric(dys_lvl), ~ max(.x, .y)
      ),
    pressure = factor(
      pressure,
      labels = pressure_labels
    )
  )
```

With the `pressure` variable we generate the variables used in the model.

```{r}
fram_long <- fram_long |>
  dplyr::mutate(
    bp_optimal = as.numeric(pressure == "optimal"),
    bp_high_normal = as.numeric(pressure == "high-normal"),
    bp_hypertension_i = as.numeric(pressure == "hypertension I"),
    bp_hypertension_ii_iv = as.numeric(pressure == "hypertension II-IV")
  )
```

## Selecting the Needed Variables

Once the Framingham variables are generated we need to create the model in the `MiceExtVal` package. The Framingham model is stratified by sex. Therefore, we need to generate two models, one for males and another for females. The `MiceExtVal` package allow us to define a Cox model using the `MiceExtVal::mv_model_cox` function where we need to introduce the model (`coefficients`), the mean value in the derivation cohort for each of the variables (`means`), the model formula to know the dependent variable (`formula`) and the survival basal function for the time of the study (`S0`).

### Generating Extra Needed Variables
Some variables like `cursmoke` or `diabetes` are factors and to be used in the model we need to transform them to numeric variables where $1$ indicates `TRUE` and $0$ indicates `FALSE`. We will also generate the variable `age2` $age^2$ needed for the females model. Additionally, we generate the survival outcome. 

```{r}
fram_long <- fram_long |>
  dplyr::mutate(
    cursmoke = as.numeric(cursmoke == "1"),
    diabetes = as.numeric(diabetes == "1"),
    age2 = age**2,
    anychd_surv = survival::Surv(timechd, anychd == "Yes")
  )
```

Finally, we generate the males and females datasets.

```{r}
male_fram_long <- fram_long |> dplyr::filter(sex == 1)
female_fram_long <- fram_long |> dplyr::filter(sex == 2)
```

### Generate the models

The models in the package are created using the `MiceExtVal::mv_model_cox` function if we want to externally validate a Cox model or `MiceExtVal::mv_model_logreg` if we want to validate a logistic regression model. In this case we want to validate the Framingham model that is a Cox model.

Each model needs different information in order to calculate the predictions in the Cox model we need to introduce the following parameters.

* `formula`: A formula that defines the model. In this formula we need to define the model coefficients and if needed, like in this case, the means to center the variables in the $\beta \cdot X$ calculation. 
* `S0`: The value of $S_0(t)$ for the time of study $t$ in the model. In this case $t = 10 \text{ years}$.

As the Framingham model is stratified by sex, we need to generate two models to externally validate it.

#### Males
The next code snippet generates the males model formula following the original mean values and the coefficients described in the previous tables.

```{r}
males_formula <- anychd_surv ~ 0.0483 * (age - 48.3) +
  0.4284 * (diabetes - 0.05) + 0.5234 * (cursmoke - 0.403) -
  0.6595 * (totchol_less_160 - 0.075) + 0.1769 * (totchol_200_239 - 0.39) +
  0.5054 * (totchol_240_279 - 0.165) + 0.6571 * (totchol_greater_280 - 0.057) +
  0.4974 * (hdl_less_35 - 0.192) + 0.2431 * (hdl_35_44 - 0.357) -
  0.0511 * (hdl_50_59 - 0.19) - 0.4866 * (hdl_greater_60 - 0.106) -
  0.0023 * (bp_optimal - 0.202) + 0.2832 * (bp_high_normal - 0.202) +
  0.5217 * (bp_hypertension_i - 0.225) +
  0.6186 * (bp_hypertension_ii_iv - 0.128)
```

Once we have defined the needed parameters of the model we can create the model as follows.

```{r}
male_model <- MiceExtVal::mv_model_cox(
  formula = males_formula,
  S0 = 0.9002
)
```

#### Females
For females we can reproduce what we have already done for males.

```{r}
females_formula <- anychd_surv ~ 0.3377 * (age - 49.6) -
  0.0027 * (age2 - 2604.5) + 0.5963 * (diabetes - 0.038) +
  0.2925 * (cursmoke - 0.378) + -0.2614 * (totchol_less_160 - 0.079) +
  0.2077 * (totchol_200_239 - 0.327) + 0.2439 * (totchol_240_279 - 0.2) +
  0.5351 * (totchol_greater_280 - 0.91) + 0.8431 * (hdl_less_35 - 0.043) +
  0.378 * (hdl_35_44 - 0.149) + 0.1978 * (hdl_45_49 - 0.124) -
  0.4295 * (hdl_greater_60 - 0.407) - 0.5336 * (bp_optimal - 0.348) -
  0.0677 * (bp_high_normal - 0.15) + 0.2629 * (bp_hypertension_i - 0.186) +
  0.4657 * (bp_hypertension_ii_iv - 0.1)
```

From the previous parameters we generate the `female_model` to externally validate the Framingham female model.

```{r}
female_model <- MiceExtVal::mv_model_cox(
  formula = females_formula,
  S0 = 0.9625
)
```

# External validation

Once we have defined the Framingham model for the package we can now start to externally validate the model. The package functions are separated in two big groups, the functions that calculate predictions or results from the model definition are started by `calculate_`, and the functions that obtain external validation results are started by `get_`.

The external validation can be structured in three phases.

1. Calculate the predictions.
2. Calculate the recalibrations. 
3. Obtain the external validation results.

## Calculate the predictions

```{r}
male_model <- male_model |>
  MiceExtVal::calculate_predictions(model = _, data = male_fram_long)

female_model <- female_model |>
  MiceExtVal::calculate_predictions(model = _, data = female_fram_long)
```

As a result inside the model there are generated two tables `predictions_data` and `predictions_aggregated`. In the next table there are shown the first rows of the `predictions_data` table, where there are all the predicted values for each of the imputations.


```{r}
#| label: tbl-pred-data
#| classes: plain .table-sm
#| tbl-cap: Example of `predictions_data`
#| echo: false
head(male_model$predictions_data) |>
  kableExtra::kbl(digits = 3)
```

In this second table there are shown the first rows of `predictions_aggregated` where there are only one aggregated prediction for each patient.

```{r}
#| label: tbl-pred-agg
#| classes: plain .table-sm
#| tbl-cap: Example of `predictions_aggregated`
#| echo: false
head(male_model$predictions_aggregated) |>
  kableExtra::kbl(digits = 3)
```

## Model performance
The model performance can be separated in two. The ability of the model to discriminate high-risk patients from low-risk patients and its calibration i.e. the ability to match the real risk and predicted risk. The model C-Index is calculated using the Harrell's C-Index and the calibration ability by generating calibration plots.

### Calculating C-Index

#### C-Index in Males

```{r}
male_model <- male_model |>
  MiceExtVal::calculate_c_index(model = _, data = male_fram_long)
```

#### C-Index in Females
```{r}
female_model <- female_model |>
  MiceExtVal::calculate_c_index(model = _, data = female_fram_long)
```

### Calculating Calibration Plot

#### Calibration Plot Males
```{r, message=FALSE, warning=FALSE}
#| fig.alt: >
#|   Calibration plot for males
#| warning: false
male_model |>
  MiceExtVal::get_calibration_plot_data(
    model = _,
    data = male_fram_long,
    n_groups = 10,
    type = "predictions_aggregated"
  ) |>
  MiceExtVal::get_calibration_plot(data = _)
```

#### Calibration Plot Females
```{r, message=FALSE, warning=FALSE}
#| fig.alt: >
#|   Calibration plot for females
#| warning: false
female_model |>
  MiceExtVal::get_calibration_plot_data(
    model = _,
    data = female_fram_long,
    n_groups = 10,
    type = "predictions_aggregated"
  ) |>
  MiceExtVal::get_calibration_plot(data = _)
```

## Recalibration
```{r, message=FALSE, warning=FALSE}
#| fig.alt: >
#|   Type 1 calibration plot for males
#| warning: false
male_model <- male_model |>
  MiceExtVal::calculate_predictions_recalibrated_type_1(
    model = _, data = male_fram_long
  ) |>
  MiceExtVal::calculate_predictions_recalibrated_type_2(
    model = _, data = male_fram_long
  )

male_model |>
  MiceExtVal::get_calibration_plot_data(
    model = _,
    data = male_fram_long,
    n_groups = 10,
    type = "predictions_recal_type_1"
  ) |>
  MiceExtVal::get_calibration_plot(data = _)
```

```{r, message=FALSE, warning=FALSE}
#| fig.alt: >
#|   Type 2 calibration plot for males
male_model |>
  MiceExtVal::get_calibration_plot_data(
    model = _,
    data = male_fram_long,
    n_groups = 10,
    type = "predictions_recal_type_2"
  ) |>
  MiceExtVal::get_calibration_plot(data = _)
```
```{r, message=FALSE, warning=FALSE}
#| warning: false
#| fig.alt: >
#|   Type 1 calibration plot for females
female_model <- female_model |>
  MiceExtVal::calculate_predictions_recalibrated_type_1(
    model = _, data = female_fram_long
  ) |>
  MiceExtVal::calculate_predictions_recalibrated_type_2(
    model = _, data = female_fram_long
  )

female_model |>
  MiceExtVal::get_calibration_plot_data(
    model = _,
    data = female_fram_long,
    n_groups = 10,
    type = "predictions_recal_type_1"
  ) |>
  MiceExtVal::get_calibration_plot(data = _)
```

```{r, message=FALSE, warning=FALSE}
#| fig.alt: >
#|   Type 2 calibration plot for females
female_model |>
  MiceExtVal::get_calibration_plot_data(
    model = _,
    data = female_fram_long,
    n_groups = 10,
    type = "predictions_recal_type_2"
  ) |>
  MiceExtVal::get_calibration_plot(data = _)
```

## References
